Life in the canopy: Community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-02-07 ms/manuscript.Rmd
Update MS as per editor comments 2020-02-05

@@

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/canopylife.
Data are openly available from the Dryad repository at http://doi.org/ADD_DOI_HERE_WHEN_ASSIGNED [@Nitta2020].

@@

@@

**`r s_table("species-list")`** List of ferns from Moorea, French Polynesia included in this study

**`r s_table("trait-values")`** Trait values of ferns on Moorea, French Polynesia

**`r s_table("ancova")`** Analysis of covariance (ANCOVA) for climate variables in response to growth habit (epiphytic vs. terrestrial) with elevation as a covariate

**`r s_table("phylosig-binary-strict")`** Phylogenetic signal in binary traits, strict dataset excluding any species whose traits were scored based on taxonomy

**`r s_table("corr-evo-strict")`** Pagel's (1994) test of correlated evolution between binary traits, strict dataset excluding any species whose traits were scored based on taxonomy

**`r s_table("model-fit")`** Best-fitting models explaining community diversity metrics by climate and growth habit

**`r s_figure("corr-plot")`** Correlation plot comparing results of functional community diversity analyses using different sets of traits as input

**`r s_figure("cwsd")`** Comparison of community-weighted standard deviations (CWSD) in traits of epiphytic and terrestrial fern communities on Moorea, French Polynesia

@@

**`r figure("traits-used-image")`** Examples of traits used in this study. **(a)** Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinnae pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. **(b)** Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars \SI{1}{\milli\meter} except for glands, which is \SI{.1}{\milli\meter}. Photographs by J. H. Nitta.

**`r figure("climate")`** Selected microclimatic variables of study plots along an elevational gradient from \SI{200}{\meter} to \SI{1200}{\meter} on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from `r final_climate_survey$start` to `r final_climate_survey$end`. Vapor pressure deficit (VPD) was calculated from relative humidity and temperature as described in the methods, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at \circa{ }\SI{2}{\meter} on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval{ }< 0.05. Partially transparent colors indicate outliers excluded from the models (see Methods).

@@

**`r figure("tree")`** Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Tree obtained from @Nitta2017. Relative value of quantitative (sporophyte) traits shown by heat gradient: low values are lighter, high values are darker. Stipe length and rhizome diameter were log-transformed prior to scaling. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Values for three leaf size traits (stipe length, frond length, and frond with) were correlated, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see `r table("traits-used")`. Major epiphytic radiations identified by @Schuettpelz2009 labeled on tree in green: H, Hymenophylloideae; T, Trichomanoideae; V, Vittarioideae; A, *Asplenium*; E, *Elaphoglossum*; P, Polypodiaceae; taxonomy follows @PteridophytePhylogenyGroupI2016.

@@

**`r figure("cwm-div")`** Change in community-weighted mean (CWM) values of epiphytic and terrestrial fern communities along an elevational gradient on Moorea, French Polynesia. Epiphytic communities in green, terrestrial communities in brown. Error bars are each one standard deviation (values below zero not shown). Trendlines and \rval^2^ shown for significant relationships with elevation or the interaction of elevation and growth habit as determined by a linear model at \pval{ }< 0.05. \tval shown for significant differences between means of CWMs by growth habit as determined by a two-sided \tval-test at \pval{ }< 0.05; \*\*\* = \pval{ }< 0.001; \*\* = \pval{ }< 0.01; \* = \pval{ }< 0.05. Results for frond width were very similar to frond length and not shown.

**`r figure("comm-div")`** Functional and phylogenetic diversity of epiphytic and terrestrial fern communities along an elevational gradient on Moorea, French Polynesia. Epiphytic communities in green, terrestrial communities in brown. Response variable abbreviations as follows: MPD~phy~, standard effect size (SES) of mean phylogenetic distance; MNTD~phy~, SES of mean nearest taxon distance; MPD~func~, SES of mean functional distance; MNTD~func~, SES of mean nearest functional distance. Trendlines indicate significant relationships as determined by a linear model at \pval{ }< 0.05. Boxplots show median values (bold lines) by growth habit. Lower and upper hinges correspond to first and third quartiles, and whiskers extend to values within 1.5 × the interquartile range. Asterisks indicate statistical significance of two-sided \tval-test; \*\*\* = \pval{ }< 0.001; \*\* = \pval{ }< 0.01; \* = \pval{ }< 0.05.

@@

**Table `r table_nums("traits-used", display = "n")`** Traits used in this study

@@

**Table `r table_nums("phylosig-cont", display = "n")`** Phylogenetic signal in continuous traits

@@

**Table `r table_nums("phylosig-binary", display = "n")`** Phylogenetic signal in binary traits

@@

**Table `r table_nums("pics", display = "n")`** Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth

@@

**Table `r table_nums("corr-evo", display = "n")`** Pagel's (1994) test of correlated evolution between binary traits

>>>>

Brief heading: "Community trait assessments reveal substantial functional diversity among fern epiphytes"

Social media    
Twitter: \@joel_nitta  
Website: \textnormal{https://joelnitta.com}

Orcid IDs   
Joel H. Nitta: \textnormal{https://orcid.org/0000-0003-4719-7472}  
Charles C. Davis: \textnormal{https://orcid.org/0000-0001-8747-1101}

@@

Code to replicate all analyses, figures, and this manuscript is available at https://github.com/joelnitta/canopylife.
Data are openly available from the Dryad repository at https://doi.org/10.5061/dryad.fqz612jps [@Nitta2020].

@@

<!-- FIXME: add DOI for Dryad Nitta et al. 2020 dataset: https://doi.org/10.5061/dryad.fqz612jps -->

@@

**`r s_table("species-list")`** `r s_table_cap("species-list")`

**`r s_table("trait-values")`** `r s_table_cap("trait-values")`

**`r s_table("ancova")`** `r s_table_cap("ancova")`

**`r s_table("phylosig-binary-strict")`** `r s_table_cap("phylosig-binary-strict")`

**`r s_table("corr-evo-strict")`** `r s_table_cap("corr-evo-strict")`

**`r s_table("model-fit")`** `r s_table_cap("model-fit")`

**`r s_figure("cwsd")`** `r s_figure_cap("cwsd")`

**`r s_figure("corr-plot")`** `r s_figure_cap("corr-plot")`

@@

**`r figure("traits-used-image")`** Examples of fern traits used in this study. **(a)** Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinnae pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. **(b)** Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars \SI{1}{\milli\meter} except for glands, which is \SI{.1}{\milli\meter}. Photographs by J. H. Nitta.

**`r figure("climate")`** Selected microclimatic variables of study plots along an elevational gradient from \SI{200}{\meter} to \SI{1200}{\meter} on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from `r final_climate_survey$start` to `r final_climate_survey$end`. Vapor pressure deficit (VPD) was calculated from relative humidity and temperature as described in the methods, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at \circa{ }\SI{2}{\meter} on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval{ }< 0.05. Asterisks indicate statistical significance of linear model; \*\*\* = \pval{ }< 0.001; \* = \pval{ }< 0.05. Partially transparent colors indicate outliers excluded from the models (see Methods).

@@

**`r figure("tree")`** Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Tree obtained from @Nitta2017. Relative value of quantitative (sporophyte) traits shown by heat gradient: low values are lighter, high values are darker. Stipe length and rhizome diameter were log-transformed prior to scaling. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Values for three leaf size traits (stipe length, frond length, and frond with) were correlated, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see `r table("traits-used")`. Major epiphytic radiations identified by @Schuettpelz2009 labeled on tree in green: H, Hymenophylloideae; T, Trichomanoideae; V, Vittarioideae; A, *Asplenium*; E, *Elaphoglossum*; P, Polypodiaceae; taxonomy follows @PteridophytePhylogenyGroupI2016. Scale in millions of years (my).

@@

**`r figure("cwm-div")`** Change in community-weighted mean (CWM) values of epiphytic and terrestrial fern communities along an elevational gradient on Moorea, French Polynesia. Epiphytic communities in green, terrestrial communities in brown. Error bars are each one standard deviation (values below zero not shown). Trendlines and \rval^2^ shown for significant relationships with elevation or the interaction of elevation and growth habit as determined by a linear model at \pval{ }< 0.05. \tval shown for significant differences between means of CWMs by growth habit as determined by a two-sided \tval-test at \pval{ }< 0.05; \*\*\* = \pval{ }< 0.001; \*\* = \pval{ }< 0.01. Results for frond width were very similar to frond length and not shown. Response variable abbreviations: SLA, specific leaf area.

**`r figure("comm-div")`** Functional and phylogenetic diversity of epiphytic and terrestrial fern communities along an elevational gradient on Moorea, French Polynesia. Epiphytic communities in green, terrestrial communities in brown. Response variable abbreviations as follows: MPD~phy~, standard effect size (SES) of mean phylogenetic distance; MNTD~phy~, SES of mean nearest taxon distance; MPD~func~, SES of mean functional distance; MNTD~func~, SES of mean nearest functional distance. Trendlines indicate significant relationships as determined by a linear model at \pval{ }< 0.05. Boxplots show median values (bold lines) by growth habit. Lower and upper hinges correspond to first and third quartiles, and whiskers extend to values within 1.5 × the interquartile range. Asterisks indicate statistical significance of two-sided \tval-test; \*\*\* = \pval{ }< 0.001; \*\* = \pval{ }< 0.01.

@@

**Table `r table_nums("traits-used", display = "n")`** Fern traits used in this study

@@

**Table `r table_nums("phylosig-cont", display = "n")`** Phylogenetic signal in continuous traits of ferns from Moorea, French Polynesia

@@

**Table `r table_nums("phylosig-binary", display = "n")`** Phylogenetic signal in binary traits of ferns from Moorea, French Polynesia

@@

**Table `r table_nums("pics", display = "n")`** Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth in ferns from Moorea, French Polynesia

@@

**Table `r table_nums("corr-evo", display = "n")`** Pagel's (1994) test of correlated evolution between binary traits in ferns from Moorea, French Polynesia
